---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r libraries, include = FALSE}
library(dplyr)
library(ggplot2)
```

```{r loading data}
activity <- read.csv(unz("activity.zip", "activity.csv"))
```

## What is mean total number of steps taken per day?

```{r total number of steps taken per day}
activity %>%
    group_by(
        date
    ) %>%
    summarise(
        total = sum(steps)
    ) -> daily_activity

head(daily_activity)
```

```{r histogram_of_total_number_of_steps_taken_each_day}
ggplot(daily_activity, aes(total)) + 
    geom_histogram(binwidth = 1000, color = "cadetblue4", fill = "cadetblue3") + 
    labs(x = "Total number of steps taken everyday", y = "Count",
         title = "Distribution of total number of steps taken each day") +
    theme_minimal()
```

```{r mean and median of the total number of steps taken per day}
mean_daily_steps <- mean(daily_activity$total, na.rm = TRUE)
median_daily_steps <- median(daily_activity$total, na.rm = TRUE)
```

The average number of steps taken per day is `r sprintf("%.2f", mean_daily_steps)`, while the median is `r sprintf("%.2f", median_daily_steps)`.

## What is the average daily activity pattern?

```{r average number of steps during interval}
activity %>%
    group_by(
        interval
        ) %>%
    summarise(
        mean_steps = mean(steps, na.rm = TRUE)
    ) -> interval_activity
```

```{r time_series_plot}
ggplot(interval_activity, aes(x = interval, y = mean_steps)) + 
    geom_line(color = "firebrick4") +
    labs(x = "Interval", y = "Average number of steps", 
         title = "Average number of steps during interval") +
    theme_minimal()
```

```{r max number of steps}
max_steps_interval <- interval_activity[which.max(interval_activity$mean_steps),]$interval
```

On average, the highest number of steps was recorded during the interval numbered `r max_steps_interval`.

## Imputing missing values

```{r number of NA}
na_count <- sum(is.na(activity))
```

The total number of missing values is `r na_count`.

Based on the analyses conducted in the previous steps, the most reasonable approach is to impute the missing values using the average number of steps for the corresponding interval.

```{r imputation}
activity %>%
  left_join(interval_activity, by = "interval") %>%
  mutate(steps = ifelse(is.na(steps), mean_steps, steps)) %>%
  select(-mean_steps) -> activity_imputed
```

```{r total number of steps taken per day after imputation}
activity_imputed %>%
    group_by(
        date
    ) %>%
    summarise(
        total = sum(steps)
    ) -> imputed_daily_activity

head(imputed_daily_activity)
```


```{r histogram_of_total_number_of_steps_taken_everyday_after_imputation}
ggplot(imputed_daily_activity, aes(total)) + 
    geom_histogram(binwidth = 1000, color = "cadetblue4", fill = "cadetblue3") + 
    labs(x = "Total number of steps taken everyday", y = "Count",
         title = "Distribution of total number of steps taken everyday after imputation") +
    theme_minimal()
```

```{r mean and median of the total number of steps taken per day after imputation}
mean_daily_steps_imputed <- mean(imputed_daily_activity$total, na.rm = TRUE)
median_daily_steps_imputed <- median(imputed_daily_activity$total, na.rm = TRUE)
```

After data imputation, the mean is `r sprintf("%.2f", mean_daily_steps_imputed)` and the median is `r sprintf("%.2f", median_daily_steps_imputed)`.

## Are there differences in activity patterns between weekdays and weekends?

```{r set local time to English, include = FALSE}
old_locale <- Sys.getlocale("LC_TIME")

Sys.setlocale("LC_TIME", "C")
```


```{r add weekday information}
activity_imputed %>%
    mutate(
        date = as.Date(date),
        weekday = weekdays(date),
        day_type = factor(ifelse(weekday %in% c("Saturday", "Sunday"),
                             "weekend", "weekday"))
    ) -> activity_imputed_weekdays

```

```{r average number of steps during interval divided by day_type}
activity_imputed_weekdays %>%
    group_by(
        day_type,
        interval
        ) %>%
    summarise(
        mean_steps = mean(steps, na.rm = TRUE),
        .groups = "drop"
    ) -> interval_activity_imputed_weekdays
```

```{r time_series_plot_for_weekdays_and_weekends}
ggplot(interval_activity_imputed_weekdays, aes(x = interval, y = mean_steps)) + 
    geom_line(color = "firebrick4") +
    facet_grid(day_type ~.) +
    labs(x = "Interval", y = "Average number of steps", 
         title = "Average number of steps during interval") +
    theme_bw()
```

```{r Restore previous locale time, include = FALSE}
Sys.setlocale("LC_TIME", old_locale)
```

